---
title: 'Data Analysis in R: Lecture Notes'
author: "PS239T"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    df_print: paged
    theme: flatly
    highlight: pygments
---

# 0) NOTES

Purpose of document is to clean dataset of Florida police shootings from 2009 to 2014. 

See github page for more information about dataset: https://github.com/tbtimes/florida-police-shooting-data



**********************************************************

# 1) SET UP 

## 1.1. General 

```{r}
# remove all objects
rm(list=ls())

# Unload all packages 
library(pacman)
pacman::p_unload(all)

# Add packages 
pacman::p_load(
  tidyverse, #dplyr, readr, etc.
  data.table, #fread() 
  foreign, #load data types including stata .dta files 
  magrittr, #%<>% operator
  skimr, #for summerising
  stargazer, #pretty tables 
  xtable, #pretty tables 
  knitr #pretty tables (kable)
)
```

## 1.2. Load Data 

```{r}
load("../data/fl_shooting_data_clean.RData")
```

**********************************************************

# 2) SUMMARISE DATA

What do you want to summarise? 

```{r}
# First, take a quick look at the data again to 
fl_shootings %>% glimpse()

# Now, let's choose some examples 

fl_shootings %>%
  summarise(fatality_mean = mean(fatality, na.rm=T),
            fatality_sd = sd(fatality, na.rm=T))


fl_shootings %>% 
  group_by(race) %>%
  summarise(n=round(n()/nrow(fl_shootings)*100, 1),
            fatality=mean(fatality, na.rm=T)) 


fl_shootings %>% 
  group_by(race) %>%
  summarise(n=n()/nrow(fl_shootings)*100,
            fatality=mean(fatality, na.rm=T)) 


```



**********************************************************

# 3) FUNCTIONS 

Writing functions using dplyr and other tidyverse funtions is similar to writing other functions in R. However, functions with variable names as arguments (e.g. x in the function below) can be difficult because some dplyr functions do not recognize character variable names. For functions with this issue (e.g. group_by), dplyr has an alternative version of the function with a '_' following the function name. 
```{r}
freqTable <- function(datName, x){
  datName %>% 
    group_by_(x) %>% 
    summarise(total=n(),
              pct=round(total/nrow(datName)*100,1)) %>% 
    arrange(desc(total)) %>%
    print()
}

eval(parse(text="freqTable(fl_shootings,'weapon_choices')"))
```

More generally, you can run any R code saved as a character string `t` using `eval(parse(text=t))`:
```{r}
freqTable <- function(datName, x){
  # Group by x
  txt <- paste0("datName %<>% group_by(",x,")") 
  eval(parse(text=txt))
  
  # Summarize and arrange data 
  datName %>% 
    summarise(total=n(),
              pct=round(total/sum(total)*100,1)) %>% 
    arrange(desc(total)) %>%
    print()
}

freqTable(fl_shootings,'weapon_choices')
```


**********************************************************

# 4) RESHAPING 

```{r}
# Frequency of weapons 
weapon_by_year <- fl_shootings %>% 
  group_by(weapon_choices, year) %>% 
  summarise(total=n()) %>%
  print() 

# Spread out totals by year 
weapon_by_year %>% 
  spread(key=year, value=total)

# Restack data 
weapon_by_year %>% 
  spread(key=year, value=total) %>%
  gather(key=year_new, value=total_new, -weapon_choices)
```

# 5) PRETTY TABLES

## 5.1. Kable 

```{r}
# Frequency of weapons 
fl_shootings %>% 
  group_by(weapon_choices) %>% 
  summarise(total=n()) %>%
  kable 
```

## 5.2. Stargazer 

```{r}
# Create example regression & print default output 
lm1 <- lm(number_of_shots ~ black + other_race, data=fl_shootings) %>% print() 

# Summary function 
lm1 %>% summary() 

# Generate and save text version of stargazer output 
lm1 %>% stargazer(type = "text", title="OLS Example", digits=3,  
                  out="../output/table1.txt")

# Generate LaTex version of stargazer output 
lm1 %>% stargazer()
```

## 5.2. xtable 

```{r}
# Frequency of weapons 
fl_shootings %>% 
  group_by(weapon_choices) %>% 
  summarise(total=n()) %>%
  xtable
```


